Synthesis of Linear Functional Programs

### 2021-07-20 article/article.tex
More restrictions paragraph

The interesting challenge of polymorphism in synthesis is the usage of
schemes from the unrestricted context.  To begin with, $\Gamma$ now
holds both types and schemes. Consequently, after the rule
\textsc{decideLeft!} is applied, we become left focused on either a
type or a scheme. Since left focus on a type is already well
formulated, only rules for left focus on a scheme are needed. Our
algorithm instances bound type vars of the focused scheme with
existential type variables, and the instanced type becomes the left
focus. Inspired by the Hindley-Milner system, we also generate
inference constraints on the existential type variables (postponing
the decision of what type it should be to be used in the proof), and
collect them in a new constraints context $\Theta$ that is propagated
across derivation branches the same way the linear context is (by
havin an input and output context ($\Theta/\Theta'$)).  In contrast to
Hindley-Milner's inference, everytime a constraint is added it is solved against all other
constraints -- a branch of the proof search is desired to fail as soon
as possible; and more, two different existential vars do not unify
against one another ($?\alpha \mapsto ?\beta$ fails) \todo{O que
  significa $?\alpha$ ? Não foi explicado -- no Hindley-Milner nao ha
  propriamente o problema de misturar variaveis existenciais com universais}. Note that we
instance the scheme with \emph{existential} type variables ($\alpha$)
rather than just type variables since the latter represent universal
types in the synth, and the first represent an existential instance of
a scheme, that might inclusively constrain normal type
variables. Additionally, it's required that all existential type
variables were assigned a type. These concepts
are formalized with the rules:
\begin{mathpar}
    \infer*[right=($\forall L$)]
    {
        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
        \\
        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
        \\
        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
    }
    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
    \and
    \infer*[right=($?L$)]
    {\textsc{unify}(?\alpha
    \mapsto C, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
    x{:}?\alpha \Downarrow\ \vdash x : C}
    \and
    \infer*[right=($\Downarrow L?$)]
    {\textsc{unify}(?\alpha
    \mapsto A, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
    x{:}A \Downarrow\ \vdash x : ?\alpha}
\end{mathpar}
where
\begin{mathpar}
    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
    \overline{\alpha}.\ \tau
    \and
    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
    \and
    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
    \textrm{to type}\ \tau_x
    \\
    \textsc{unify}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
    \textrm{can be unified with those in}\ \Theta
\end{mathpar}

>>>>

The interesting challenge of polymorphism in synthesis is the usage of
schemes from the unrestricted context.  To begin with, $\Gamma$ now
holds both types and schemes. Consequently, after the rule
\textsc{decideLeft!} is applied, we become left focused on either a
type or a scheme. Since left focus on a type is already well
formulated, only rules for left focus on a scheme are needed. Our
algorithm instances bound type variables of the focused scheme with
existential type variables, and the instanced type becomes the left
focus. Inspired by the Hindley-Milner system, we also generate
inference constraints on the existential type variables (postponing
the decision of what type it should be to be used in the proof), and
collect them in a new constraints context $\Theta$ that is propagated
across derivation branches the same way the linear context is (by
havin an input and output context ($\Theta/\Theta'$)).  In contrast to
Hindley-Milner's inference, everytime a constraint is added it is solved against all other
constraints -- a branch of the proof search is desired to fail as soon
as possible.
%TODO: Como fazer? \todo{O que
% significa $?\alpha$ ? Não foi explicado -- no Hindley-Milner nao ha
% propriamente o problema de misturar variaveis existenciais com universais}
Note that we
instance the scheme with \emph{existential} type variables ($?\alpha$)
rather than just type variables ($\alpha$) since the latter represent universal
types in the synth, and the first represent an existential instance of
a scheme, that might inclusively constrain traditional type
variables. Additionally, it's required that all existential type
variables are assigned a type. These concepts
are formalized with the rules:
\begin{mathpar}
    \infer*[right=($\forall L$)]
    {
        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
        \\
        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
        \\
        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
    }
    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
    \and
    \infer*[right=($?L$)]
    {\textsc{unify}(?\alpha
    \mapsto C, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
    x{:}?\alpha \Downarrow\ \vdash x : C}
    \and
    \infer*[right=($\Downarrow L?$)]
    {\textsc{unify}(?\alpha
    \mapsto A, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
    x{:}A \Downarrow\ \vdash x : ?\alpha}
    \\
    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
    \overline{\alpha}.\ \tau
    \and
    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
    \and
    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
    \textrm{to type}\ \tau_x
    \\
    \textsc{unify}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
    \textrm{can be unified with those in}\ \Theta
\end{mathpar}

\mypara{More Restrictions} We can now look at two more sources of infinite
recursion in the synthesis process. The first is the use of an unrestricted
function to synthetise a term of type $\tau$ that in turn will require a term of
the same type $\tau$. A good example is the sub-goal judgment $\Gamma,(a \lolli
b \lolli b); (a \lolli b \lolli b) \Downarrow\ \vdash b$ that
appears while synthetising \emph{foldr} -- we apply ($\lolli$L)
until we can use \textsc{init} ($b \Downarrow\ \vdash b$), and then we must
synthetise an argument of type $b$. Without any additional restrictions, we
can end up left focused on $(a \lolli b \lolli b)$ again, and again require $b$,
and on and on. The solution will be to disallow the usage of the same function
to synthetise the same goal a second time further down in the derivation.

% TODO TODO: Big rewrite
The other situation occurs when using an unrestricted polymorphic function that
requires synthesis of a term with an existential type when the goal is an
existential type. In contrast to the previous problem, the type of the goal and
of the argument that will cause the loop won't be the same, since instantiated
bound variables are always fresh. For example, for $\Gamma,\forall \alpha,\beta
.\ \alpha\lolli\beta\lolli\beta;?\alpha\lolli?\beta\lolli?\beta \Downarrow\
\vdash\ ?\sigma$, we'll unify $?\beta$ with $?\sigma$, and then require a term
of type $?\beta$ (not $?\sigma$). We want to forbid the usage of the \emph{same}
function to attain \emph{any} existential goal, provided that function might
create existential sub-goals (i.e. it's polymorphic). However, we noticed that,
even though for most tried problems this ``same function'' approach worked,
context-heavy problems such as \emph{array} (seen in Section~\ref{sec:overview})
wouldn't terminate in a reasonable amount of time.  In fact, with $n$
polymorphic functions in the unrestricted context, the complexity of searching for a
program of any existential type $?\alpha$, while restricting solely the used
function, is $O(n!)$~\ref{exemplo_apendice?}. As such, we'll instead forbid the
usage of \emph{all} unrestricted polymorphic functions to synthetise any
existential type if, altogether, we've already used $d_e$ times a polymorphic
function for \emph{any} existential sub-goal, where $d_e$ is a \emph{constant
existential depth} that controls the maximum depth of ``existential synthetis''.
This approach reduces the complexity of our proof-search algorithm for
existential types to $O(\tfrac{n!}{d_e!}) = O(n^{d_e})$.

Extending the restrictions context ($\Rho$) with restrictions on using the
unrestricted context ($\Rho_{L!}$), we modify \textsc{decideLeft!} to formalize
the previous paragraph:
\begin{mathpar}
    \mprset{flushleft}
    \infer*[right=(decideL!)]
    {
    (A, C) \notin \Rho_{L!}
    \\
    \textsc{isExist}(C) \Rightarrow |\{u\ |\
    (f,u)\in\Rho_{L!},\textsc{isExist}(u),\textsc{isPoly}(f)\}| < d_e
    \\
    \Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!}');\Gamma, A; \Delta/\Delta' ; A
    \Downarrow\ \vdash C
    }
    {\Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!});\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
    \\
    \Rho_{L!}' =\ \textbf{if}\ \textsc{isFunction}(A)\ \textbf{then}\ \Rho_{L!},(A, C)\ \textbf{else}\ \Rho_{L!}
\end{mathpar}
